---
title: 'Batlinks: occurence data from NBN'
output:
  html_document: default
  html_notebook: default
  word_document: default
params:
  path.occ:
    label: "Path to occurrence data:"
    value: "data/observations/Chiroptera_NBN_since1990"
  fn.occ:
    label: "File name: include file extension (e.g. .csv).  Leave blank if you are reading in a whole folder of files."
    value: "Chriroptera_NBN_since1990.csv"
  sel.taxon.level:
    label: "Select minimum taxon level"
    value: "species"
    input: select
    choices: [species, genus, family, suborder, order, unranked]
  sel.coord.uncert:
    label: "Select maximum coordiante uncertainty (metres)"
    value: "1000"
    input: select
    choices: [1, 10, 50, 100, 1000, 5000, 10000]
  sel.min.occ:
    label: "Select minimum number of records per species"
    value: "1"
---
   

```{r setup, include=FALSE}
#document setup
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, echo=FALSE)
```



```{r, include=FALSE}
#packages-----
library(skimr)
library(knitr)
library(kableExtra)
library(sp)
library(rgdal)
library(tidyverse)
library(rnrfa)
```


```{r}
#variables

#occurrence data to import
#path to raw occurrence data
path.occ <- paste0("../", params$path.occ)  

#filename raw occurrence data
fn.occ <- params$fn.occ

#data selection criteria
sel.taxon.level <- params$sel.taxon.level #taxon level

#create list of taxon levels
#species, genus, family, suborder, order, unranked
taxa <- c("species", "genus", "family", "suborder", "order", "unranked")
sel.taxon.level <- taxa[1:match(params$sel.taxon.level, taxa)] #creates vector of taxon level and below. 

sel.coord.uncert <- params$sel.coord.uncert #coordinate uncertainty in metres
sel.min.occ <- params$sel.min.occ #minimum number of occurrences per taxon

```


```{r}
#functions

#table formatting for output
format_table <- function(table){
  knitr::kable(table, caption = table_caption) %>% 
  kableExtra::kable_styling(bootstrap_options = "striped", 
                            full_width = F, 
                            position = "left")
}
```

This document summarises bat occurence data for the Batlinks area that is available publically on NBN Atlas.


```{r}
#load data

#read single occurrence file
# occ_NBN_raw <- read_csv(paste0(path.occ, fn.occ))
# occ_NBN <- occ_NBN_raw

#read folder of occurrence files (all in same format)
#speciesdf <- map_df(list.files(path = path.occ, full.names = TRUE), read_csv)
#read single occurrence file
speciesdf <- read_csv(file = paste0(path.occ, "/", fn.occ))
names(speciesdf)
head(speciesdf)
```


```{r}
#remove duplicates
paste("Raw data has ", nrow(speciesdf), "records")
speciesdf <- distinct(speciesdf) 
paste("After removal of duplicates there are now ", nrow(speciesdf), "records.")

#Prepare the data for analysis by adding eastings and northings
#code adapted from github.com/jncc/sdms

## Data from the NBNatlas Inital clean to give presence records in GB, also subset records by bngCol as varying precisions for Grid References
speciesdf$OSGR <- toupper(speciesdf$OSGR)
speciesdf <- speciesdf[speciesdf$`Occurrence status` == "present",]  # remove absence data
#speciesdf <- speciesdf[!speciesdf$`State/Province` == "Northern Ireland",]  # subset to GB only
speciesdf <- speciesdf[grepl("[[:alnum:]]", speciesdf[["OSGR"]]),]  #select rows with gridref records
names(speciesdf)[names(speciesdf) == "Year"] <- "year"
speciesdf$precision <- NA  #add precision

# ensure grid references are consistent
speciesdf$precision <-
  ifelse(grepl("^[a-zA-Z]{2}[0-9]{4}$", speciesdf[["OSGR"]]), '1km', 'NA')
  speciesdf$precision[grepl("^[a-zA-Z]{2}[0-9]{2}[a-zA-Z]{1}$", speciesdf[["OSGR"]])] <-
  '2km'
  speciesdf$precision[grepl("^[a-zA-Z]{2}[0-9]{2}$", speciesdf[["OSGR"]])] <-
  '10km'
  speciesdf$precision[grepl("^[a-zA-Z]{2}$", speciesdf[["OSGR"]])] <-
  '100km'
  speciesdf$precision[grepl("^[a-zA-Z]{2}[0-9]{6}$", speciesdf[["OSGR"]])] <-
  '100m'

    # Check precision column is numerical, and convert to m
    if (!is.numeric(speciesdf[["precision"]])) {
        speciesdf$res <- sub("[^[:alpha:]]+", "", speciesdf[["precision"]])
        speciesdf[["precision"]] <- as.numeric(gsub("([0-9]+).*$", "\\1",
            speciesdf[["precision"]]))
        speciesdf[["precision"]] <- ifelse(speciesdf$res == "m", speciesdf[["precision"]],
            speciesdf[["precision"]] * 1000)
        speciesdf$res <- NULL
    }  
  
  
paste(nrow(speciesdf), "occurrences after subsetting")  
```


```{r}
#create easting and northing columns
speciesdf$easting<-NA
speciesdf$northing<-NA

# Split between normal grid refs and 'tetrad' (2km) grids
nontetrad <- speciesdf[which(!speciesdf[["precision"]] == 2000),]
tetrad <- speciesdf[which(speciesdf[["precision"]] == 2000),]

nontetrad[c("easting", "northing")] <- nontetrad$OSGR %>% 
      purrr::map(
        purrr::possibly(osg_parse, 
                        otherwise = list(easting= NA, northing = NA), 
                        quiet = T)
      ) %>% 
  do.call(rbind.data.frame,.)

select(nontetrad, OSGR, easting, northing)
```


```{r}
#this loop fails with incorrect grid refs, so use purrr function above
# if (nrow(nontetrad) > 0) {
#   for (i in 1:nrow(nontetrad)) {
#     ne <- rnrfa::osg_parse(nontetrad$OSGR[i])
#     nontetrad$easting[i] <- ne[[1]]
#     nontetrad$northing[i] <- ne[[2]]
#   }
# }

speciesdf <- nontetrad

    # Calculate tetrad grids seperately
    if (nrow(tetrad) > 0) {
        tetrad$At10kcorner <- paste(stringi::stri_sub(tetrad[[bngCol]],
            1, 4))
        tetrad$Letter <- paste(stringi::stri_sub(tetrad[[bngCol]], 5,
            -1))
        for (i in 1:nrow(tetrad)) {
            ne <- rnrfa::osg_parse(tetrad$At10kcorner[i])
            tetrad$easting[i] <- ne[[1]]
            tetrad$northing[i] <- ne[[2]]
        }
        Tetrad <- c("A", "B", "C", "D", "E", "F", "G", "H", "I", "J",
            "K", "L", "M", "N", "P", "Q", "R", "S", "T", "U", "V", "W",
            "X", "Y", "Z")
        AddEast <- c(1000, 1000, 1000, 1000, 1000, 3000, 3000, 3000, 3000,
            3000, 5000, 5000, 5000, 5000, 5000, 7000, 7000, 7000, 7000,
            7000, 9000, 9000, 9000, 9000, 9000)
        AddNorth <- c(1000, 3000, 5000, 7000, 9000, 1000, 3000, 5000,
            7000, 9000, 1000, 3000, 5000, 7000, 9000, 1000, 3000, 5000,
            7000, 9000, 1000, 3000, 5000, 7000, 9000)
        AddTetrad <- data.frame(Tetrad, AddEast, AddNorth)
        tetrad <- merge(tetrad, AddTetrad, by.x = "Letter", by.y = "Tetrad",
            all.x = T)
        tetrad$easting <- tetrad$easting + tetrad$AddEast
        tetrad$northing <- tetrad$northing + tetrad$AddNorth
        tetrad$At10kcorner <- tetrad$AddEast <- tetrad$AddNorth <- tetrad$Letter <- NULL
        colnames(tetrad)[1] <- colnames(nontetrad)[1]
        speciesdf <- rbind(nontetrad, tetrad)

    }  

#write to a clean object
occ_NBN <- speciesdf
```


```{r}
#summarise and save to object
summary_occ_NBN <- skim(occ_NBN)

#store parsing problems
probs_occ_NBN <- problems(occ_NBN)

#select actual values from problem cells
probs_occ_NBN <- cbind(probs_occ_NBN, value = occ_NBN[probs_occ_NBN$row, unique(probs_occ_NBN$col)])  
```
## Data 

```{r}
table_caption <- "Parsing problems"
probs_occ_NBN[-5] %>% 
  format_table()
```



```{r}
##Prepare data

#select variables
#names(occ_NBN)
#occ_NBN <- occ_NBN_raw
occ_NBN <- occ_NBN %>% 
  select(
    # TCI = `Taxon Concept ID`,
    TCI = `Species ID (TVK)`,
    # taxon_rank = `Taxon rank`,
    spp_name = `Scientific name`,
    # order = Order,
    # genus = Genus,
    # species = Species,
    #datum = `Geodetic Datum`,
    gridref = `OSGR`,
    lat = `Latitude (WGS84)`, 
    lon = `Longitude (WGS84)`,
    easting = easting, 
    northing = northing,
    coord_uncert = precision,
    date = `Start date`
    #year = year#,
    #month = Month
    ) 

#Make the occurrence data spatial
#count(occ_NBN, datum)

#remove records without location
occ_NBN <- filter(occ_NBN, !is.na(lon))

occ_NBN_coords <- select(occ_NBN, lon, lat)
  
occ_NBN_sp <- SpatialPointsDataFrame(coords = occ_NBN_coords, data = occ_NBN,
                                     proj4string = CRS("+proj=longlat +datum=WGS84"))

#have a look at the projections
#proj4string(occ_NBN_sp); proj4string(bound); proj4string(bbox)

#transform projection to match polygons
occ_NBN_sp <- spTransform(occ_NBN_sp, CRSobj = CRS(proj4string(bound)))
#plot(occ_NBN_sp)
```

```{r}
#plot boundaries and points
# sp::plot(bound, main = "Project boundary and location of records")
# #plot(bbox)#, add = T)
# points(occ_NBN_sp, add=T)

ggplot() +  
  geom_point(data = occ_NBN_sp %>% 
               coordinates() %>% 
               data.frame() %>% 
               mutate(species = occ_NBN_sp$spp_name), 
             aes(x=lon, y=lat, colour = species)) +
  geom_polygon(data = bound, aes(x=long, y=lat), 
               fill = "green", colour = "black", alpha = 0.01) +
  labs(title = "Project boundary and location of records", 
       x = "easting", 
       y = "northing") 
```

```{r}
#select points within boundary 
occ_NBN_sp <- occ_NBN_sp[bound,]
plot(occ_NBN_sp)
```

```{r}
#write selected data back to a dataframe
occ_NBN <- occ_NBN_sp@data
```



```{r}
#Sample of data (random selection of rows)
table_caption <- "Sample of data (random selection of rows)"
occ_NBN %>% 
  sample_n(20) %>% 
  format_table()
```

## Taxa


```{r}
table_caption <- "Count of species"
occ_NBN %>% 
  count(spp_name, sort = T) %>% 
  format_table()
```


Consider location uncertainty. 

```{r}
table_caption <- "Coordinate uncertainty in metres"
graphdata <- occ_NBN %>% 
  select(spp_name, coord_uncert) %>% 
    mutate(species = as.factor(spp_name)) 

for(i in unique(graphdata$species)){
print(ggplot(filter(graphdata, species == i), aes( x = coord_uncert)) +
  geom_bar() +
  scale_x_log10()+
    labs(title = i)) 
}
```

```{r, message=FALSE, warning=FALSE}
occ_table <- occ_NBN %>% 
  filter(spp_name != is.na(spp_name) ) %>% 
  select(spp_name, coord_uncert) %>%
  mutate(species = as.factor(spp_name)) %>% 
  mutate(coord_uncert = as.factor(coord_uncert)) %>% 
  group_by(species) %>% 
  count(coord_uncert) %>% 
  spread(key = coord_uncert, value = n)

occ_table

write.csv(occ_table, "../data/occ_table.csv", row.names = F)


# 
#  group_by(species) %>% 
#   count(species, sort = T) %>% 
#   summarise_if
```


coordinate uncertainty by species
```{r}
table_caption <- "Coordinate uncertainty in metres by species"
occ_NBN %>% 
  group_by(spp_name) %>% 
count(coord_uncert) %>% 
  format_table()
```

## Select data for further analysis

Select records that are: 

* identified to `r params$sel.taxon.level` level
* coordinate uncertainty no greater than `r sel.coord.uncert`m
* at least `r sel.min.occ` records per species 

**This isn't working properly except for taxon level species.  I think that's because I am selecting at different taxon levels, but then filtering by the column `species`.  Should instead use `scientific name interpreted`. **

```{r}
occ_sel <- occ_NBN %>% 
  filter(
   coord_uncert <= sel.coord.uncert
         ) 

sppcount <- occ_sel %>% 
  count(spp_name) %>% 
  filter(n >= sel.min.occ) 

occ_sel <- occ_sel %>% 
  filter(spp_name %in% sppcount$spp_name)

#tabulate changes
table_caption <- "Number of records following selection"
full_join(x = count(occ_sel, spp_name), 
          y = count(occ_NBN, spp_name), 
          by = "spp_name", 
          suffix = c(".sel", ".orig")) %>% 
  mutate(diff = n.sel - n.orig) %>% 
  format_table()
```




```{r}
#Make the occurrence data spatial

occ_sel_coords <- select(occ_sel, lon, lat)
  
occ_sel_sp <- SpatialPointsDataFrame(coords = occ_sel_coords, data = occ_sel, proj4string = CRS("+proj=longlat +datum=WGS84"))

#have a look at the projections
#proj4string(occ_sel_sp); proj4string(bound); proj4string(bbox)

#transform projection to match polygons
occ_sel_sp <- spTransform(occ_sel_sp, CRSobj = CRS(proj4string(bound)))
#plot(occ_sel_sp)

#sp::plot(bound, main = "Project boundary and location of records")
#plot(bbox)#, add = T)
#points(occ_sel_sp, add=T)
```


```{r}
#prepare data for ggplot plotting
ggplot() +  
  geom_polygon(data = bound, aes(x=long, y=lat), 
               fill = "white", colour = "black", alpha = 0.01) +
  geom_point(data = occ_sel_sp %>% 
               coordinates() %>% 
               data.frame() %>% 
               mutate(spp_name = occ_sel_sp$spp_name), 
             aes(x=lon, y=lat, colour = spp_name)) +
  labs(title = "Project boundary and location of records", 
       x = "easting", 
       y = "northing")

```

```{r}
table_caption <- "Number of records by species"
count(occ_sel, spp_name) %>% 
  format_table()
```


```{r}
#prepare output for maxent
occ_maxent <- select(occ_sel, spp_name, easting, northing)
```


## Export data
```{r, echo=TRUE}
write_csv(speciesdf, "../data/observations/prepped/occurrence_NBN_raw.csv")
write_csv(occ_NBN, "../data/observations/prepped/occurrence_NBN_fltrd.csv")
write_csv(occ_sel, "../data/observations/occurrence_NBN_slctd.csv")
write_csv(occ_maxent, "../data/observations/occurrence_maxent.csv")

```

